---
title: "Exploratory Analysis for HW2"
author: "Takeen Shamloo"
editor: source
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    embed-resources: true
---

## Exploring Historical Production and Delivery Data for California Water Suppliers

## Dataset Information

**Source:** Historical water data from the California Data Water Consortium.  

**Variables:**  

- `pwsid`: Public Water System ID number associated with water system. Can be linked with `org_id`.
- `water_system_name`: Name of the water system.
- `org_id`: Organization ID assigned to the urban water supplier.
- `start_date`, `end_date`: Reporting period covering one month.
- `water_produced_or_delivered`: Categorizes data as either "water produced" or "water delivered".
- `water_type`: Type of water (e.g., treated, raw, recycled, groundwater, surface water, etc.).
- `quantity_acre_feet`: Converted quantity of water in acre-feet.

**Questions to Answer:**  
1. Are there seasonal or regional patterns in production rates or delivery delays?  
2. Which regions experience the highest delays, and why?  
3. How do production rates correlate with delivery efficiency?  

**Data Access:** CSV file available via the California Natural Resources Agency portal.  

#### Load Packages

```{r}
#| warning: FALSE
library(tidyverse)
library(ggplot2)
library(here)
library(dplyr)
```

### Load Data

```{r}
#| warning: FALSE
# Load data
historical_df <- read_csv(here("data", "historical_production_delivery.csv"))
```

## Data Wrangling

```{r}
# Clean and transform data
historical_df_clean <- historical_df %>%
  filter(!is.na(quantity_acre_feet)) %>%
  mutate(start_date = as.Date(start_date),
         end_date = as.Date(end_date),
         quantity_million_af = quantity_acre_feet / 1e6) # Scale to million acre-feet
```

## Exploratory Data Visualizations

### Plot 1: Water Delivered by Selected Water Types

```{r}
filtered_types <- historical_df_clean %>%
  filter(water_type %in% c("industrial", "landscape irrigation", "recycled", "groundwater wells"))

ggplot(filtered_types, aes(x = water_type, y = quantity_million_af, fill = water_type)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Sum of Water Delivered by Selected Water Types", 
    x = "Water Type", 
    y = "Quantity (Million Acre-Feet)",
    fill = "Water Types") +
  theme_minimal() +
  coord_flip()
```

### Plot 2: Seasonal Trends in Water Production (Filtered & Scaled to Million Acre-Feet)

```{r}
ggplot(filtered_types, aes(x = start_date, y = quantity_million_af, color = water_type)) +
  geom_line() +
  labs(
    title = "Seasonal Trends in Water Production", 
    x = "Date", 
    y = "Quantity (Million Acre-Feet)",
    fill = "Water Types") +
  theme(
    legend.position = "right",
    legend.justification = "top") 
```

### Plot 3: Regions with the Highest Delays
```{r}
delay_analysis_df <- historical_df_clean %>%
  group_by(org_id, water_type) %>%
  summarize(avg_quantity = mean(quantity_million_af, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(avg_quantity)) %>%
  slice_max(avg_quantity, n = 10)

ggplot(delay_analysis_df, aes(x = reorder(org_id, avg_quantity), y = avg_quantity, fill = water_type)) +
  geom_col() +
  coord_flip() +
  labs(title = "Regions with the Highest Delays in Water Delivery", x = "Organization ID", y = "Average Quantity (Million Acre-Feet)") +
  scale_fill_viridis_d() +
  theme_minimal()
```

## Reflection

### **1. What have we learned about this data?**

**Seasonality is clear:** Water production and delivery fluctuate across the year, with some months seeing significantly higher or lower delivery rates (**Plot 2**).

**Regional disparities exist:** Certain regions consistently **deliver more water** than others, and the delays in delivery appear concentrated in specific areas (**Plot 3**).

**Delivery efficiency depends on water type:** Some types (like **industrial and recycled water**) have more consistent production, while others fluctuate more (**Plot 1**).

### **2. How are we doing on our original questions?**

**Are there seasonal/regional patterns?** → **Yes**, production and delivery vary **seasonally** (**Plot 2**) and **regionally** (**Plot 3**).

**Which regions experience the highest delays, and why?** → We identified the **top 10 regions** with the highest delivery delays (**Plot 3**), but need more analysis to determine the reasons (e.g., infrastructure, policies).

**How do production rates correlate with efficiency?** → Not fully answered yet—**next steps** should include comparing production and delivery rates over time.

### **3. What challenges are ahead?**

**Time-series alignment:** Making sure that data is properly **aggregated and compared across months** so trends remain accurate.

**Understanding delays:** We need more **external data** (e.g., policy or infrastructure info) to determine **why** some regions have delays.

**Further correlation analysis:** Investigating whether **higher production rates lead to greater efficiency or more delays**.
